The effect of climate forcing on numerical simulations of the Cordilleran ice sheet at the Last Glacial Maximum

### 2014-03-31 cordillera-climate.tex
Negate offsets and lapse rate in figs. and text

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of 6\,\unit{{\degree}C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from 0 to 10\,\unit{{\degree}C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a~5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

@@

To contrast the effect of the six different input climatologies, we plot simulations using a~5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

@@

Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of 6\,\unit{{\degree}C\,km^{-1}} as in the simulations. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, which may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}). However significant disparity among reanalyses exists, too. More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a ``hybrid'' climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of 5\,\unit{{\degree}C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim 5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

@@

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8\,\unit{{\degree}C} (WorldClim), 6\,\unit{{\degree}C} (ERA-Interim) and 7\,\unit{{\degree}C} (NARR). These qualitative ``best'' runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of 7 to 11\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.

@@

% fig:cool05
\begin{figure*}
  \includegraphics{cordillera-climate-cool05}
  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate 5\,\unit{{\degree}C} colder than present for each climate forcing used.}
  \label{fig:cool05}
\end{figure*}

% fig:tempheatmap
\begin{figure}
  \includegraphics{cordillera-climate-tempheatmap}
  \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis. Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of 6\,\unit{{\degree}C\,km^{-1}}. Colour mapping is based on a~logarithmic scale. Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
  \label{fig:tempheatmap}
\end{figure}

@@

% fig:biatm
\begin{figure*}
  \includegraphics{cordillera-climate-biatm}
  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using ``hybrid'' climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
  \label{fig:biatm}
\end{figure*}

% fig:biatmbars
\begin{figure}
  \includegraphics{cordillera-climate-biatmbars}
  \caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim 5\,\unit{{\degree}C} offset simulation. Corresponding ice sheet geometries are presented in Figs.~\ref{fig:cool05} and~\ref{fig:biatm}.}
  \label{fig:biatmbars}
\end{figure}

@@

% fig:durationstack
\begin{figure}
  \includegraphics{cordillera-climate-durationstack}
  \caption{Left panel: model sensitivity to simulation length. Modelled glaciated area, using NARR forcing data, and temperature offsets from 0 to 15\,\unit{{\degree}C}, solid lines corresponding to values from 7 to 11\,\unit{{\degree}C}. Right panel: modelled ice margin corresponding to temperature offsets from 7 to 11\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading). Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
  \label{fig:durationstack}
\end{figure}

>>>>

Prior to surface mass balance computation, the model dynamically applies a~lapse-rate correction to surface air temperature. This correction accounts for the evolution of ice thickness on the one hand, and differences between the climate forcing reference topography and the ice flow model basal topography on the other hand. It uses a~reference topography distinct from the modelled basal topography and specific to each climate forcing dataset (Fig.~\ref{fig:topo}). An annual lapse rate of -6\,\unit{{\degree}C\,km^{-1}} is used in all simulations. No corrections apply to precipitation changes with elevation.

@@

Using WorldClim, NCEP/NCAR, ERA-Interim, CFSR (smoothed and original datasets) and NARR input climatologies as climate forcing, we run 66~simulations of glacial inception and growth of the Cordilleran ice sheet over 10\,kyr using temperature offsets ranging from -10 to 0\,\unit{{\degree}C}. Significant differences in patterns of ice sheet growth, final glacial extent, and volume are attained (Figs.~\ref{fig:ivolarea}--\ref{fig:cool05}).

Across the range of temperature offsets used, final glaciated area and final ice volume differ widely between different input climatologies (Fig.~\ref{fig:ivolarea}). For instance, using a~-5\,\unit{{\degree}C} temperature offset, final glaciated area differs between WorldClim and CFSR forcing by a~factor of~6 and final volume by a~factor of~12 (Fig.~\ref{fig:ivolarea}). For particular temperature offset values, given input climatologies result in similar final glaciated areas and ice volumes, yet regional patterns are different (Fig.~\ref{fig:extent}).

@@

To contrast the effect of the six different input climatologies, we plot simulations using a~-5\,\unit{{\degree}C} temperature offset (Fig.~\ref{fig:cool05}). The magnitude of glaciation differs dramatically from one climate forcing to another. Whereas the NCEP/NCAR and CFSR climatologies produce a~large ice sheet that covers most of the model domain (Fig.~\ref{fig:cool05}, lower panels), WorldClim, ERA-Interim and NARR climatologies lead to a~more restrictive ice cover (Fig.~\ref{fig:cool05}, upper panels). The WorldClim run only resulted in the growth of mountain ice caps bounded to the highest topography.

@@

Because most of the disparity between reanalytic and observational temperature data is caused by topographical detail at scales unresolved in the reanalysis datasets, we apply a lapse-rate correction prior to this comparison, using the same lapse rate of -6\,\unit{{\degree}C\,km^{-1}} as in the simulations. ERA-Interim, CFSR and NARR climatologies exhibit temperature distributions fairly consistent with WorldClim data (Fig.~\ref{fig:tempheatmap}). The spatial distribution of temperature differences between reanalyses and WorldClim data shows partly consistent patterns, which may relate to weaknesses of WorldClim data in areas of sparse observational coverage (Fig.~\ref{fig:tempdiff}). However significant disparity among reanalyses exists, too. More specifically, surface air temperature data from the NCEP/NCAR climatology shows the largest disparity with that of WorldClim, with a~significant cold anomaly over most of the modelling domain (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}).

@@

To distinguish the effects of temperature and precipitation biases on the ice sheet model, we run a~series of eight additional simulations. These new simulations use a ``hybrid'' climate forcing that consists of temperature data from WorldClim, combined with precipitation data from each of the four reanalyses, and vice-versa. A~single temperature offset of -5\,\unit{{\degree}C} is used. This value was chosen to allow for comparison with previous results (Fig.~\ref{fig:cool05}). In each case, temperature lapse rates are computed using the reference topography from the corresponding dataset.

Although only temperature or precipitation changes are applied, this experiment results once more in large differences in modelled ice sheet geometries (Fig.~\ref{fig:biatm}) as compared to the reference WorldClim -5\,\unit{{\degree}C} run (Fig.~\ref{fig:cool05}, top-left panel). For ERA-Interim, CFSR and NARR climatologies, precipitation anomalies clearly dominate the differences in ice sheet response, whereas temperature anomalies have relatively little effect (Fig.~\ref{fig:biatmbars}). However in the case of NCAR forcing, both the negative temperature bias (Figs.~\ref{fig:tempheatmap} and \ref{fig:tempdiff}) and the important precipitation differences (Figs.~\ref{fig:precheatmap} and \ref{fig:precdiff}) contribute approximately equally to produce oversized ice sheets (Fig.~\ref{fig:biatmbars}).

@@

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Figs.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we restricted our analysis to the WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a~similar size, we select for each climate forcing the simulation that leads to a~final glaciated area closest to the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$. This corresponds to temperature offset values of -8\,\unit{{\degree}C} (WorldClim), -6\,\unit{{\degree}C} (ERA-Interim) and -7\,\unit{{\degree}C} (NARR). These qualitative ``best'' runs are presented in Fig.~\ref{fig:best} along with their associated temperature depressions and a~reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

To test the effect of an inception period of shorter duration on model results, we use simulations run under NARR climate forcing with temperature offsets of -11 to -7\,\unit{{\degree}C}, and compare modelled ice sheet extent when glaciated area reaches the approximate size of the LGM Cordilleran ice sheet of $2\times 10^6\,\unit{km^2}$.

@@

% fig:cool05
\begin{figure*}
  \includegraphics{cordillera-climate-cool05}
  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr under a~climate -5\,\unit{{\degree}C} colder than present for each climate forcing used.}
  \label{fig:cool05}
\end{figure*}

% fig:tempheatmap
\begin{figure}
  \includegraphics{cordillera-climate-tempheatmap}
  \caption{Density maps showing a~comparison of summer (JJA) surface air temperature data from the WorldClim climatology, against that of each reanalysis. Climate data are presented after bilinear spatial interpolation and correction for topographic differences to WorldClim data, using a lapse rate of -6\,\unit{{\degree}C\,km^{-1}}. Colour mapping is based on a~logarithmic scale. Note the cold bias of NCEP/NCAR data relative to WorldClim data.}
  \label{fig:tempheatmap}
\end{figure}

@@

% fig:biatm
\begin{figure*}
  \includegraphics{cordillera-climate-biatm}
  \caption{Ice surface topography (1\,\unit{km} contours) and velocity (\unit{m\,yr^{-1}}) after 10\,kyr using ``hybrid'' climate forcing with precipitation rate from WorldClim and surface air temperature from each reanalysis (upper row), and surface air temperature from WorldClim and precipitation rate from each reanalysis (lower row). In other words, the upper row shows the effect of temperature anomalies, and the lower row the effect of precipitation anomalies, for each reanalysis, relative to WorldClim data. Each simulation uses a~-5\,\unit{{\degree}C} offset for comparison with Fig.~\ref{fig:cool05}.}
  \label{fig:biatm}
\end{figure*}

% fig:biatmbars
\begin{figure}
  \includegraphics{cordillera-climate-biatmbars}
  \caption{Effect of temperature and precipitation anomalies (separately and jointly) from each reanalysis on modelled final ice volume relative to the result of the WorldClim -5\,\unit{{\degree}C} offset simulation. Corresponding ice sheet geometries are presented in Figs.~\ref{fig:cool05} and~\ref{fig:biatm}.}
  \label{fig:biatmbars}
\end{figure}

@@

% fig:durationstack
\begin{figure}
  \includegraphics{cordillera-climate-durationstack}
  \caption{Left panel: model sensitivity to simulation length. Modelled glaciated area, using NARR forcing data, and temperature offsets from -15 to 0\,\unit{{\degree}C}, solid lines corresponding to values from -11 to -7\,\unit{{\degree}C}. Right panel: modelled ice margin corresponding to temperature offsets from -11 to -7\,\unit{{\degree}C} when total glaciated area reach the approximate size of the LGM Cordilleran ice sheet of $2 \times 10^6\,\unit{km^2}$. A~reconstructed LGM ice sheet margin by \citet{dyke-2004} (Fig.~\ref{fig:locmap}) is given for reference (grey shading). Note that shorter (and colder) simulations lead to more restrictive glaciation of the continental eastern margin but further ice extent in the maritime south-western part of the modelling domain.}
  \label{fig:durationstack}
\end{figure}
